---
title: "Naming things"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("learnr")
```

## Assigning a value to a name

If you run an R command, the result will be printed to the screen.

```{r}
mean(1:10) # mean of sequence of numbers from 1:10
# result printed to screen
```

If you want to use the value again, you need to assign it to a name so it can be stored.

The usual assign operator is `<-`. The name is put on the left of the arrow and the value on the right.
The result can be accessed 

```{r}
results <- mean(1:10) # mean of sequence of numbers from 1:10
# result printed to screen
```



## Legal names

### Allowed characters

Object names can only include

- lower case letters
- UPPER case letters
- numbers
- dots `.` and underscores `_`

The first character must be a letter or a dot.

```{r legal_names_quiz, echo = FALSE}
question_checkbox(
  "Select all the names that are legal",
  answer("tomato", correct = TRUE),
  answer("_apple", correct = FALSE, message = "Names cannot start with an underscore"),
  answer("BASil", correct = TRUE),
  answer("extra virgin olive oil", correct = FALSE, message = "Names cannot include spaces"),
  answer(".banana", message = "The object `.banana` will be invisible", correct = TRUE),
  answer("2onions", correct = FALSE, message = "Names cannot start with a number"),
  answer("mushroom2", correct = TRUE),
  answer("carrot/cabbage", correct = FALSE, message = "Names can only include letters, numbers, `.` and `_`"),
  answer(".", correct = TRUE, message = "The object `.` will be invisible"),
  answer("spinach_potato", correct = TRUE),
  answer("kål", correct = TRUE, message = "Names with `æ`, `ø`, or `å` are legal but can cause huge problems with encodings. Best avoided."),
  random_answer_order = TRUE,
  allow_retry = TRUE,
  try_again = "Be sure to select all seven legal names!"
)
```

### Invisible objects

If the first character in a name is a dot, the object is invisible (this is sometime useful). 
Use `ls(all.names = TRUE)` to find any invisible objects. 

```{r invisible_names_quiz, echo = FALSE}
question_checkbox(
  "Select all statements that are true",
  answer("Object names that start with a `.` are invisible", correct = TRUE),
  answer("Object names cannot start with a `.`", correct = FALSE),
  answer("Invisible objects appear in the RStudio Environment tab", correct = FALSE),
  answer("The R command `ls(all.names = TRUE)` will show invisible objects", correct = TRUE),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

### Special characters

Object names can include characters with diacritics such as `æ`, `ø`, and `å` but these are best avoided. 

 - You may collaborate with people who do not have these characters on their keyboard.
 - There are several characters that look like `å` but are different differently by R.
 - Representing these characters depends on the file encoding. 
File encoding problems open up a whole world of pain you will want to avoid!

```{r invisible_names_quiz, echo = FALSE}
question_checkbox(
  "Select all statements that are true",
  answer("Object names can include characters with diacritics such as `ø` and `å`.", correct = TRUE),
  answer("Using characters with diacritics is a bad idea.", correct = TRUE),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

## Reserved names

Some names are reserved for R.
These include `function`, `if`, `else`, `for`, `while`, `next` and `TRUE`. 
You can see the full list with `?Reserved`.
If you try to use one of these names, you will get an error.

```{r reserved_names1, error = TRUE}
function <- 7 #error
```
```{r reserved_names2, error = TRUE}
next <- 1 #error
```
```{r reserved_names3, error = TRUE}
TRUE <- FALSE #error!
```


```{r reserved_names_quiz, echo = FALSE}
question_checkbox(
  "Select all the legal names",
  answer("TRUE", correct = FALSE),
  answer("false", correct = TRUE),
  answer("function", correct = FALSE),
  answer("NULL", correct = FALSE),
  answer("na", correct = TRUE),
  answer("mean", correct = TRUE, message = "It is legal but can be confusing to give objects the same name existing R functions such as `mean`"),
  answer("Bergen", correct = TRUE),
  answer("Function", correct = TRUE),
  random_answer_order = TRUE,
  allow_retry = FALSE,
  try_again = "Be sure to select all five legal names!"
)
```

## Case Sensitivity

R is a case sensitive language. 
This means that `fun`, `Fun` and `FUN` are all different objects.

## Good names

    “There are only two hard things in Computer Science: cache invalidation and naming things.”

    — Phil Karlton

- Names should be legal
- Names should be concise
- Names should be meaningful

Code completion in RStudio can help 

## Combining words

Make the word breaks clear

- snake_case
- camelCase
- UpperCamelCase
- best.avoided (possible confusion with R fuctions)