---
title: "Data handling with Tidyverse for beginners"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("flair")
```

## Introduction to the `tidyverse`

The `tidyverse` package is set of R packages for data handling and display.
I use it for all my data handling.

You can install `tidyverse` with 

```{r installation, eval = FALSE}
install.packages("tidyverse")
```

This can take a few minutes. 
You only need to do this once.

Load `tidyverse` with

```{r load-tidyverse, include = FALSE}
library("tidyverse")
```

Some important packages in `tidyverse` are

- **`ggplot2`** for graphics
- **`tibble`** for improved `data.frame`s
- **`readr`** for importing text files (e.g. csv files)
- `readxl` for importing excel files
- **`tidyr`** for tidying data
- **`dplyr`** for manipulating `data.frame`s or `tibble`s
- **`stringr`** for manipulating character vectors
- `lubridate` for manipulating dates (optionally with times) and `hms` for times
- **`forcats`** for manipulating factors
- `broom` for extracting summary information from models
- **`purrr`** for iterating 

Packages in bold are automatically loaded with `tidyverse`, others need loading with `library()` separately. 
This tutorial focuses on `tidyr` and `dplyr`. 


## Pipes

It is very common to need to run functions on the output of other functions. 
For example, the following code uses `pull` to extract the `Sepal.Width` column from the `iris` data set, then calculates the variance, and then square roots this to give the standard deviation.

```{r without-pipes}
x <- sqrt(var(pull(iris, Sepal.Width), na.rm = TRUE))
x
```

It is fairly dificult to read isn't it. 
It can be difficult to see which argument belongs to each function: sometime an argument gets given to the wrong function.

An alternative way to write this code would be to break it into seperate steps

```{r intermediates}
x <- pull(iris, Sepal.Width)
x2 <- var(x, na.rm = TRUE)
x3 <- sqrt(x2)
x3
```

This is easier to read, but we end up with a lot of intermediate objects and keeping track of them, for example if we wanted to change the order of the functions, is difficult.

This is where pipes can be used to chain functions together. The pipe operator is `%>%` (the Rstudio keyboard shortcut is `ctrl-shift-m`). 
Pipes are available when `tidyverse` is loaded. 

```{r flair, echo = FALSE}
decorate("with_pipe") %>% 
  flair("%>%")
```


```{r with_pipe, include = FALSE}
x <- iris %>%
  pull(Sepal.Width) %>% 
  var(na.rm = TRUE) %>% 
  sqrt()
x
```

The pipe works by passing the result of the function on the left of the pipe to the first argument of the function after the pipe.

If we don't want to pass the result to the first argument, we either need to name the first argument (the result will go to the first available argument), or safer and more explict, use the `.` pronoun to specify where the result should go. 

```{r, error=TRUE}
iris %>% lm(Sepal.Length ~ Sepal.Width) # Fails. The data are given to the first argument which should be the formula
iris %>% lm(formula = Sepal.Length ~ Sepal.Width) # Works. Data are given to first available argument. Confusing
iris %>% lm(Sepal.Length ~ Sepal.Width, data = .) # Better using . pronoun
```

There are some other types of pipe in the `magrittr` package that can be useful occasionally.

## tibbles and data frames

A `tibble` is an improved version of a `data.frame` for storing data in columns.

  The main advantages are with printing. Unless a `data.frame` is very large, R will print it all. In contrast, a printing a `tibble` will only show what fits on the screen. The `tibble` also shows us what type of object is in each column. 

```{r tibbles, results='markup'}
# a data.frame
iris

# a tibble
as_tibble(iris)
```

Importing data with the `readr` or `readxl` packages will automatically make a `tibble`. The function `as_tibble` will convert a `data.frame` into a `tibble` (if you want to keep rownames you will need to use the rownames argument or pass the `data.frame` to `rownames_to_column()` function first)

All functions that work with a `data.frame` should work with a `tibble` (it is probably a bug if it doesn't). 


## Making tidy data with `tidyr`
       
pivot_longer pivot_wider

(complete crossing drop_na expand extract fill full_seq nesting replace_na separate separate_rows table1 uncount unite) 
      
The `tidyr` cheatsheet is very useful.

## dplyr

`dplyr` is a package for manipulating `data.frames` and `tibbles`.
There are many function in `dplyr`, this focuses on the functions I use most often.

This tutorial assumes that you have `dplyr` version >= 1.0.0. To find your `dplyr` version, run `packageVersion("dplyr")`. 

__THIS VERION IS CURRENTLY ONLY AVAILABLE ON GITHUB `remotes::install_github("tidyverse/dplyr")` and is called 0.8.xx__

The `dplyr` cheatsheet is very useful.

## Selecting columns

You can choose which columns of the `data.frame` you want with `select()`.

The first argument is the data, which is supplied by the pipe `%>%`, the next arguments are the names of the columns you want. 
The names do not need quote marks.

```{r select1}
#select Sepal.Length, Sepal.Width, Species
iris %>% select(Sepal.Length, Sepal.Width, Species)
```

This is equivalent to the base R code

```{r select1.5}
#select Sepal.Length, Sepal.Width, Species
iris[, c("Sepal.Length", "Sepal.Width", "Species")]
```


If you want to `select` adjacent columns, you can use the notation `first:last`.

```{r select2}
#select Sepal.Length to Petal.Width
iris %>% select(Sepal.Length:Petal.Width)
```

Sometimes it is easier to remove the columns you don't want. 
You can do this by putting a `-` in front of the column name.

```{r select3}
#select Sepal.Length, Sepal.Width, Species
iris %>% select(-Petal.Length, -Petal.Width)
```

If there is a pattern to the column names that we want to select (or remove), there are some helper functions.
For example, to select columns that start with "Sepal", we can use the function `starts_with`.

```{r select4}
#select Sepal.Length, Sepal.Width, Species
iris %>% select(starts_with("Sepal"))
```

Conversly, if we want to select all columns that end with "Width", we can use the fuction `ends_with()`.
`contains()` is more flexible and `matches()` is the most powerful of the helper functions, using regular expressions to identify the columns.

Sometimes, you might want to select all the columns of a certain type. 
For example, to select all the numeric columns we can use the `is.numeric` function inside select.

```{r select5}
iris %>% select(is.numeric) # No brackets on the function
```

Other `is.*` functions exist, for example, `is.character` for text, `lubridate::is.Date` for dates.

You can also select columns by number (1 being the first column), but this is generally a bad idea because if a new column is added, the code will break.

## Renaming columns

You can use `rename()` to rename columns

```{r rename1}
iris %>% select(spp = Species)
```

The syntax is `new_name = current_name`.

You can also rename a column when selecting

```{r rename2}
iris %>% select(spp = Species)
```


## Moving columns

Sometimes it is useful to reorder the columns. 
This is never necessary for data analysis or plotting, but can be needed when making a table for presentation.

```{r relocate}
iris %>% relocate(Species)
```

The defaut is to move the named column first, the `.before` and `.after` arguments let you move the column into any position. 

## filtering rows

Filtering rows that meet some condition is a very common task.

For example, to filter rows that have a sepal length greater than 6, we can use 

```{r filter1}
iris %>% filter(Sepal.Length > 6)
```

This will filter out each row where the condition is TRUE.

Other mathematical tests include 
- `==` exactly equals. Remember two equals signs for a test of equality. A single equals sign is used to assign a value to an argument. Sometimes it is safer to use the function `near` which has a built-in tolerance of rounding errors

```{r filter-near}
sqrt(2) ^ 2 == 2 # should be true, but rounding errors 
sqrt(2) ^ 2 - 2 # the difference
near(sqrt(2) ^ 2, 2) # safe alternative
```

- `!=` not equal to
- `<` less than
- `<=` less than or equal to
- `>` greater than
- `>=` greater than or equal to


The base R equivalent of this is 

```{r filter_base}
iris[iris$Sepal.Length > 6, ]
```

Generally, `filter` makes it easier to understand, especially with more complex criteria.

If we want to filter on multiple criteria, we need to decide whether we want all criteria to be TRUE (AND in Boolean logic), or for one or more to be TRUE (OR in Boolean logic).

If we want all criteria to be TRUE, we can separate them by a comma (or by an `&` if you want to be explicit).

```{r filter-AND}
iris %>% filter(Sepal.Length > 6, Petal.Length > 6)
```

If we want any criteria to be TRUE, we can separate them by a `|`.

```{r filter-OR}
iris %>% filter(Sepal.Length > 6 | Petal.Length > 6)
```

We can negate a criterion by putting `!` in front of it.
So to filter rows that do not have sepals longer than 6 we can use

```{r filter-NOT}
iris %>% filter(!Sepal.Length > 6)
```

Of course, in this example, we could also use `<=` as the test.

Other useful conditions include

- `is.na()` for filtering by missing values.
- `between()` for filtering values with a range
- `near()` a safer alterative to `==` as `==`
- `%in%` is used when you want to test if a value is in a vector

```{r filter-in}
iris %>% filter(Species %in% c("setosa",     "versicolor"))
#equivalent to 
iris %>% filter(Species == "setosa" | Species == "versicolor")
```

## Slicing the data

Sometimes it is useful to extract rows by number. 

```{r slice}
iris %>% slice(3:7)
```

You can use negative numbers to remove rows. Be careful using `slice()` as is the data change, different rows may be returned. 


## Distinct

If there are duplicates...

## Random rows

Sometimes you want to sample rows at random from a data.frame. 
Two useful functions for this are `sample_n` and `sample_frac`. 
The first samples a constant _n_ rows, the second samples a constant fraction of the rows.

## Mutating and adding columns


mutate (transmute)
lag lead recode n_distinct  n coalesce last first nth row_number 
cummean cumall cumany

case_when if_else

## summarise

## grouping data
group_by ungroup
 -> summarise
 -> mutate
 
## count
count (tally)


(intersect setdiff union union_all)


((group_map group_modify group_walk))

## sorting data
arrange desc 


## multi-table
left_join inner_join full_join (right_join) ((semi_join anti_join))
bind_cols bind_rows 

## General problems

### Non standard names

Ideally column names should follow the standard rules for naming objects in R - UPPER and lower case letters, numbers, "." and "_" with the first character being a letter (or a dot if you want an invisible object).
If you have a column name that doesn't follow these rules, you need to enclose it with back-ticks.

```{r non-standard-names}

```

It is sometimes best to rename these columns


## advanced ??
nest unnest
map

## arrange_all arrange_at arrange_if filter_all filter_at filter_if distinct_all distinct_at distinct_if group_by_all group_by_at group_by_if mutate_all mutate_at mutate_if rename_all rename_at rename_if select_all select_at select_if transmute_all transmute_at transmute_if summarise_all summarise_at summarise_if vars funs 