---
title: "Processing text with `stringr`"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 1, fig.width = 5)
library("learnr")
library("tidyverse")
theme_set(theme_classic())

tutorial_options(exercise.cap = "Exercise")
```

Consider this vector of diatom species names (which could be a column in a `data.frame` or `tibble`)

```{r}
diatoms <- c("Navicula_elkab",
             "Nitzschia_palaea",
             "Aulacoseira_granulata") # a vector
```

We might want to detect which elements of the vector have a particular pattern, or replace, remove or extract part of the text. We can do this with the `stringr` package which is part of `tidyverse`.

```{r eval = FALSE}
library("tidyverse")# load stringr and ggplot, dplyr etc
```

This tutorial starts with detecting or replacing fixed patterns and then shows how you can use regular expressions to extract varying patterns.

## Detecting a pattern

We might want to detect which of the species are in the genus _Navicula_. We can do this with `str_detect` (the base R equivalent is `grepl`.

```{r}
str_detect(string = diatoms, pattern = "Navicula")
```
This return a logical vector, `TRUE` where diatoms includes the pattern "Navicula", `FALSE` otherwise.

If the vector `diatoms` was a column in a `tibble` (or `data.frame`), we can use this test in a `filter` to select rows. 

```{r}
diatom_df <- tibble(species = diatoms, 
                    count = c(27, 3, 46))
diatom_df

diatom_df %>% 
  filter(str_detect(string = diatoms, pattern = "Navicula"))
```

## Replacing text

We can replace characters in some text using `str_replace`. So to replace the underscore in diatoms with a space, we could use

```{r}
str_replace(diatoms, pattern = "_", replacement = " ")
```

This will replace the first underscore in each element. If there were several underscores and we want to replace them all, we can use `str_replace_all`. If we want to remove some character, we can either use `str_replace` and set `replacement = ""`, or use `str_remove`.

We can use `str_replace` on a column in a tibble with a `mutate`

```{r}
diatom_df %>% 
  mutate(species = str_replace(string = species,
                               pattern = "_", 
                               replacement  = " "))
```

## Regular expressions

Sometimes what you want to detect, replace or extract is not a fixed pattern. In this case, regular expressions are a very powerful tool. For a fuller a

https://regex101.com/ is a great 

### Choices



### Repeating sequences

We can control how many times something gets matched by following it with a quantifier. 

- Zero or more times `?`

```{r}
#match both palaeoecology and paleoecology
str_detect(c("palaeoecology", "paleoecology"), pattern = "pala?eo")
```


- One or more times `+`

- Any number of times `*`

- Exactly twice {2}

```{r}
#extract year from code
str_extract("x2020", pattern = "\\d{4}")
```


- Between two and four times {2, 4}

- At most four times {, 4}

- At least twice {2, }







anchors
^$

number of repeats
? + * {}

.
\\d \w \s

choice
[a-z] 
[aeiou]
|
```{r}
str_extract_all("Navicula elkab", "\\w*")
```

capture groups
()

Can also use regular expression in `matches` to select columns of a tibble 

```{r}
str_trim()
str_sub()
str_wrap("The quick brown fox jumps over the lazy dog.", width = 30)

```


excaping

fixed()
